\textsc{\wobble: a data-driven method for precision radial velocities}

### 2018-11-28 paper/paper.tex
reverting changes

\author[0000-0002-9328-5652]{Daniel Foreman-Mackey}
\affiliation{\flatiron}

@@

\author[0000-0002-0296-3826]{Rodrigo Luger}
\affiliation{\flatiron}

@@

Our approach is similar to that taken by \citet{Gao2016}, but differs in that we take the stellar spectrum as a latent variable to be optimized rather than iteratively determining it by stacking model residuals.
%In this sense, ours is a true forward model.  -- this content-less (DFM)

@@

In Section \ref{s:methods}, we outline the model and its key underlying assumptions.
We present an open-source implementation of this method in \python and \TF
\citep{Abadi15} called \wobble.
In Section \ref{s:results}, we demonstrate \wobble's capabilities by applying our method to \HARPS archival data for three target stars: the planet-hosting solar analog 51 Peg, the quiet M dwarf \Mdwarf, and the Hot Jupiter host HD 189733.
We look further into the detailed time-variable telluric spectra inferred from these data in Section \ref{s:tellurics}.
We revisit many of the assumptions underlying \wobble in Section \ref{s:future} with a detailed look at the limitations of the current implementation.
We outline potential ways of adapting \wobble\ for such cases as instruments with absorption cells, intrinsic time variability in the stellar spectrum, and lower-quality data.
Finally, we conclude with a brief summary in Section \ref{s:conclusion}.

@@

We assume that the spectra can be modeled as the product of a finite and fixed number of components.
For the cases shown in this work, two components are used: a stellar spectrum which is invariant in shape but may be Doppler-shifted, and a telluric spectrum which is fixed to the observatory rest frame but varies in shape.
We choose to work in log(flux) space so that the data to be modeled are simply a sum of the component spectra.

@@

Aside from this regularization to push model components to zero in the low-\SNR regime, the model makes no assumptions about the shape of the stellar or telluric spectra.
We solve for spectral templates for each component as a series of control points with no imposed correlations between them, meaning that the line spread function, covariances between lines arising from the same species, and other such physically expected correlations are not built into the model but must be learned in the process of optimizing.
In addition to keeping this model simple, linear, and largely convex
(\todo{MB -- what does ``largely'' mean here?}), this means that no physical knowledge about the object being observed is needed to extract its \RVs.

For practical applications, the number of epochs $N$ must be large (\todo{MB
-- how large? 12? 50? 100? Probably good to give some sense}) and span a significant fraction of the observing season.
If the number of spectra is small or if all spectra are taken over a very small range of time (e.g. a single night), it will be nearly impossible to disentangle the components of the model, as this process relies on the apparent Doppler shifting of the star(s) and Earth's atmosphere relative to each other.
We will revisit this restriction in Section \ref{s:future}.

@@

We model each data column $y_n$ as the sum of the stellar and telluric spectra:
\begin{equation}
y_n = y_{n}^\star + y_{n}^t + noise.
\end{equation}

The stellar spectrum contribution is:
\begin{equation}
y_{n}^\star = P(v_{obs, n}) \mu_{\star},
\end{equation}
where $\mu_{\star}$ is a spectral template with length $M'$ corresponding to
the length of an arbitrarily-chosen wavelength grid of control points, and
$P$ is a linear operator which applies a Doppler shift by velocity $v_n$ and
interpolates $\mu_\star$ into the observatory rest frame.
Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
\begin{equation}
 \ln \lambda(v) = \ln \lambda_{0} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
\end{equation}
The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space.
\todo{MB -- feel free to cut this:} $P$ could be represented as a rectangular
$M \times M'$ sparse matrix, but the dot product can be implemented
efficiently using linear interpolation instead of instantiating the full matrix.

@@

The telluric spectrum contribution is:
\begin{equation}
y_{n}^t =  A_n(\mu_{t} + W_{t} z_n).
\end{equation}
In addition to its mean spectrum $\mu_t$, the telluric component also includes a time-dependent term assembled from two variables: $W_t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$.
$W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of basis vectors used.
For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications.
The net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $A_n$, a known quantity.

We evaluate a log-likelihood for each epoch
\begin{equation}
\ln \mathcal{L}_n = -\frac{1}{2} (y_{n} - y'_{n})^T C_{n}^{-1} (y_{n}-y'_{n}),
\end{equation}
comparing our model $y'$ to the data $y$, with $C$ representing the covariance matrix of uncertainties for the data.

@@

L1 normalization adds a term to the log-likelihood that takes the form \todo{(TO DO: change notation to match literature)}:
$$ L1(p, \lambda) = -\lambda \sum_{i} | p_{i} | ,$$
where $p$ is the vector of parameters to be normalized (in this case $\mu_{{\star}}$, $\mu_{{t}}$, or $W_{t}$) and $\lambda$ is the regularization amplitude.

Similarly, L2 normalization adds a term of the form:
$$ L2(p, \lambda) = -\lambda \sum_{i} p_{i}^2 .$$

@@

We set regularization amplitudes for \wobble using a cross-validation scheme.
In brief, we randomly select 12.5\% of the total epochs to set aside as a validation set and, using some value of $\lambda$, run the model optimization on the remaining epochs (the training set).
The resulting best-fit spectral templates and basis vectors are taken as fixed and the time-dependent terms only (\RVs and basis weights) are optimized for the validation epochs.
The $\chi^2$ for the validation epochs can then be adopted as a goodness-of-fit measurement for the $\lambda$ value, and the procedure is repeated for different $\lambda$s to choose the best regularization amplitude.
Because we have multiple regularization terms and multiple amplitudes to set, we begin by hand-setting all amplitudes to a reasonable starting guess and optimize each amplitude sequentially with cross-validation.

@@

After convergence, we estimate the uncertainties on the parameters by
approximating the likelihood function near its maximum as Gaussian where the
covariance matrix is the negative inverse Hessian (or second derivative
matrix) of the log likelihood function with respect to the parameters.

@@

\software{%
    astropy \citep{Astropy13, Astropy18},
    numpy \citep{Van-Der-Walt:2011},
    matplotlib \citep{Hunter:2007},
    scipy \citep{Jones:2001},
    tensorflow \citep{Abadi15}
}

>>>>

\author{Daniel Foreman-Mackey}
\affiliation{\flatiron}

@@

\author{Rodrigo Luger}
\affiliation{\flatiron}

@@

Our approach is similar to that taken by \citet{Gao2016}, but differs in that we take the stellar spectrum as a latent variable to be optimized rather than iteratively determining it by stacking model residuals. 
In this sense, ours is a true forward model.

@@

In Section \ref{s:methods}, we outline the model and its key underlying assumptions. 
We present an open-source implementation of this method in \python and \TF called \wobble. 
In Section \ref{s:results}, we demonstrate \wobble's capabilities by applying our method to \HARPS archival data for three target stars: the planet-hosting solar analog 51 Peg, the quiet M dwarf \Mdwarf, and the Hot Jupiter host HD 189733. 
We look further into the detailed time-variable telluric spectra inferred from these data in Section \ref{s:tellurics}.
We revisit many of the assumptions underlying \wobble in Section \ref{s:future} with a detailed look at the limitations of the current implementation. 
We outline potential ways of adapting \wobble\ for such cases as instruments with absorption cells, intrinsic time variability in the stellar spectrum, and lower-quality data. 
Finally, we conclude with a brief summary in Section \ref{s:conclusion}.

@@

We assume that the spectra can be modeled as the product of a finite number of components. 
For the cases shown in this work, two components are used: a stellar spectrum which is invariant in shape but may be Doppler-shifted, and a telluric spectrum which is fixed to the observatory rest frame but varies in shape. 
We choose to work in log(flux) space so that the data to be modeled are simply a sum of the component spectra.

@@

Aside from this regularization to push model components to zero in the low-\SNR regime, the model makes no assumptions about the shape of the stellar or telluric spectra. 
We solve for spectral templates for each component as a series of control points with no imposed correlations between them, meaning that the line spread function, covariances between lines arising from the same species, and other such physically expected correlations are not built into the model but must be learned in the process of optimizing. 
In addition to keeping this model simple, linear, and largely convex, this means that no physical knowledge about the object being observed is needed to extract its \RVs.

For practical applications, the number of epochs $N$ must be large and span a significant fraction of the observing season. 
If the number of spectra is small or if all spectra are taken over a very small range of time (e.g. a single night), it will be nearly impossible to disentangle the components of the model, as this process relies on the apparent Doppler shifting of the star(s) and Earth's atmosphere relative to each other. 
We will revisit this restriction in Section \ref{s:future}.

@@

We model each data column $y_n$ as the sum of the stellar and telluric spectra:
\begin{equation}
y_n = y_{\star, n} + y_{t, n} + noise.
\end{equation}

The stellar spectrum contribution is: 
\begin{equation}
y_{\star, n} = P(v_{obs, n}) \mu_{\star},
\end{equation}
where $P$ is an interpolation operator which applies a Doppler shift by velocity $v_n$ and $\mu_{\star}$ is a spectral template with length $M'$ corresponding to the length of an arbitrarily-chosen wavelength grid of control points. 
Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
\begin{equation}
 \ln \lambda(v) = \ln \lambda_{0} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
\end{equation}
The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space.

@@

The telluric spectrum contribution is:
\begin{equation}
y_{t, n} =  A_n(\mu_{t} + W_{t} z_n).
\end{equation}
In addition to its mean spectrum $\mu_t$, the telluric component also includes a time-dependent term assembled from two variables: $W_t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$. 
$W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of basis vectors used. 
For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications. 
The net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $A_n$, a known quantity.

We evaluate a log-likelihood for each epoch
\begin{equation}
\ln \mathcal{L}_n = \sum_{m} -0.5 (y_{m,n} - y'_{m,n})^T C_{m,n}^{-1} (y_{m,n}-y'_{m,n}),
\end{equation}
comparing our model $y'$ to the data $y$, with $C$ representing the covariance matrix of uncertainties for the data.

@@

L1 normalization adds a term to the log-likelihood that takes the form \todo{(TO DO: change notation to match literature)}:
$$ L1(p, \lambda) = \lambda \sum_{i} | p_{i} | ,$$
where $p$ is the vector of parameters to be normalized (in this case $\mu_{{\star}}$, $\mu_{{t}}$, or $W_{t}$) and $\lambda$ is the regularization amplitude.

Similarly, L2 normalization adds a term of the form:
$$ L2(p, \lambda) = \lambda \sum_{i} p_{i}^2 .$$

@@

We set regularization amplitudes for \wobble using a cross-validation scheme. 
In brief, we set aside 12.5\% of the total epochs as a validation set and, using some value of $\lambda$, run the model optimization on the remaining epochs (the training set). 
The resulting best-fit spectral templates and basis vectors are taken as fixed and the time-dependent terms only (\RVs and basis weights) are optimized for the validation epochs. 
The $\chi^2$ for the validation epochs can then be adopted as a goodness-of-fit measurement for the $\lambda$ value, and the procedure is repeated for different $\lambda$s to choose the best regularization amplitude. 
Because we have multiple regularization terms and multiple amplitudes to set, we begin by hand-setting all amplitudes to a reasonable starting guess and optimize each amplitude sequentially with cross-validation.

@@

The inverse variance uncertainty on the parameters is calculated as the Hessian, or the second derivative of the likelihood with respect to each parameter. \todo{(DFM: help me rephrase if needed?)}

@@

\software{\todo{TO DO! --} astropy, numpy, matplotlib, scipy, tensorflow}
